---
title: "Group 2 Project Report"
format: html
editor: visual
author: Danni Cox, Sophia Hemsi, Sarah Mulligan
---

## 1.1a Use a discrete density-dependent and -independent population model to assess the population growth dynamics of any three species within the 10 most abundant species.

```{r}

library('tidyverse')
library('tidyr')
library('dplyr')
library('DescTools')
library('vegan')
library('ggplot2')
library('lubridate')
library('ecotraj')

data <- read_csv('group2.csv')

# Identify top 3 species

species_abundance <- data |>
  group_by(species) |>
  summarise(total_abundance = sum(n, na.rm = T)) |>
  arrange(desc(total_abundance)) |>
  slice(1:10)

top3species <- species_abundance$species[1:3]

top3species

data$lambda = NA

# Plot top 3 species over time

data_over_time <- data |>
  filter(species %in% top3species) |>
  mutate(date = as.Date(date, format = "%m/%d/%Y"),
         year = year(date)) |> 
  group_by(species, year) |>
  summarise(N = sum(n, na.rm = T),
            avg_N = n(),
            .groups = "drop") |>
  arrange(species, year) |>
  ungroup()

ggplot(data_over_time, aes(x = year, y = N, color = species)) +
  geom_point() +
  geom_line() +
  labs(title = "Abundance of Top 3 Species Over Time",
       x = "Year",
       y = "Abundance",
       color = "Species") +
 
  theme_bw()
```

```{r}

# Density-independent model

data_w_lambda <- data_over_time |>
 arrange(species, year) |>
   mutate(lambda = lead(N)/N)

avg_lambda <- data_w_lambda |>
  group_by(species) |>
  summarise(
    mean_lambda = Gmean(lambda, na.rm = T),
    mean_sd = Gsd(lambda, na.rm = T)
  )

avg_lambda

# Plot lambda over time

ggplot(data_w_lambda |> filter(!is.na(lambda)),
       aes(x = year, y = lambda, color = species)) +
  geom_point() +
  geom_line() +
  labs(title = "Lambda Over Time", x = "Year", y = "Lambda", color = "Species") +
  theme_bw()

trends <- avg_lambda |>
  mutate(pop_trends = case_when(
    mean_lambda > 1.0 ~ "Growing",
    mean_lambda == 1.0 ~ "Satble",
    mean_lambda < 1.0 ~ "Declining"
  ))

trends


```

```{r}

# Project the population (density-independent model)

years <- 100

pop <- data_w_lambda |>
  group_by(species) |>
  slice_max(year, n = 1) |>
  select(species, N_0 = N, t_0 = year) |>
  left_join(avg_lambda, by = "species")

sim <- tibble()

for(i in 1:nrow(pop)) {
  Species <- pop$species[i]
  N0 <- pop$N_0[i]
  lambda <- pop$mean_lambda[i]
  t0 <- as.Date(pop$t_0[i])   
  
  N <- numeric(years + 1)
  N[1] <- N0
  
  for(t in 1:years) {
    N[t + 1] <- lambda * N[t]
  }
  
  projected_pop <- tibble(
    species = Species,
    year = 0:years,
    projection = N,
    t_i = t0 + years(0:years)  
  )
  
  sim <- bind_rows(sim, projected_pop)
}

sim

ggplot(sim, aes(x = year, y = projection, color = species)) +
  geom_line() +
  facet_wrap(~species, scales = "free_y") +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Abundance of Top 3 Species Using a\nDensity-Independent Population Model",
       x = "Years", y = "Projected Abundance", color = "Species") +
  theme_bw()
```

```{r}

# Estimate K for density-dependent model

pop2 <- data_w_lambda |>
  group_by(species) |>
  slice_max(year, n = 1) |>
  select(species, N_0 = N, t_0 = year) |>
  left_join(avg_lambda, by = "species") |>
  mutate(rd = mean_lambda - 1)

K_estimates <- tibble()

for(sp in top3species) {
  species_data <- data_over_time |>
    filter(species == sp) |>
    arrange(year) |>
    mutate(
      Nt = lead(N),
      change_in_N = (Nt - N) / N) |>
        filter(!is.na(change_in_N), is.finite(change_in_N))
      
      model <- lm(change_in_N ~ N, data = species_data)
      
      intercept <- coef(model)[1]
      
      slope <- coef(model)[2]
      
      if(slope < 0) {
        K_est <- -intercept / slope
        method <- "regression"
      } else {
        K_est <- max(species_data$N, na.rm = T) * 3
        method <- "max plus buffer"
      }
      
      K_estimates <- bind_rows(K_estimates,
                               tibble(species = sp,
                                      K = K_est,
                                      method = method))
}

# Project the population (density-dependent model)

pop2 <- pop2 |>
  left_join(K_estimates, by = "species")

pop2

sim2 <- tibble()

years <- 1000

for(i in 1:nrow(pop2)) {
  Species <- pop2$species[i]
  N0 <- pop2$N_0[i]
  rd <- pop2$rd[i]
  K <- pop2$K[i] 
  t0 <- pop2$t_0[i]   
  
  N <- numeric(years + 1)
  N[1] <- N0
  
  for(t in 1:years) {
    N[t + 1] <- N[t] + rd * N[t] * (1 - N[t] / K)
    
    if (N[t + 1] < 0) {
      N[t + 1] <- 0
    }
  }

  projected_pop2 <- tibble(
    species = Species,
    year = 0:years,
    projection = N,
    t_i = t0 + years(0:years),  
    K = K,
    rd = rd
  )
  
  sim2 <- bind_rows(sim2, projected_pop2)
}

sim2

ggplot(sim2, aes(x = year, y = projection, color = species)) +
  geom_line() +
  geom_hline(data = sim2 |> distinct(species, K),
             aes(yintercept = K, color = species),
             linetype = "dashed", linewidth = 1) +
  labs(title = "Abundance of Top 3 Species Using\na Density-Dependent Population Model",
    x = "Years",
    y = "Projected Abundance",
    color = "Species"
  ) +
    facet_wrap(~species, scales = "free_y", ncol = 1) +
  scale_y_continuous(labels = scales::comma) +
  theme_bw()

```

## 1.1b What can you infer from both models (Modeli \| Speciesi)? For each species, what is the most proper model to assess their growth? Why?

```{r}

# Summary table comparing both models

comparison_table <- pop2 |>
  select(species, N_0, rd, K, method) |>
  left_join(
    avg_lambda |> select(species, mean_lambda),
    by = "species"
  ) |>
  mutate(
    lambda_trend = case_when(
      mean_lambda > 1.0 ~ "Growing",
      mean_lambda < 1.0 ~ "Declining",
      TRUE ~ "Stable"
    ),
    density_dependent_trend = case_when(
      N_0 < K & rd > 0 ~ "Growing toward K",
      N_0 < K & rd < 0 ~ "Declining toward zero",
      N_0 > K ~ "Declining toward K",
      N_0 == K ~ "At equilibrium",
      TRUE ~ "Complex"
    )
  )

comparison_table
```

Density-independent models assume unlimited resources and constant lambda. Based on this model, Black Drum and Spotted Seatrout grew exponetially, while Gizzard Shad showed exponential decline.

The density-dependent model incorporates an estimate of carrying capacity (K), which represents resource limitations. Black Drum grew in a sigmoidal pattern, growing exponentially until nearing carrying capacity and reaching a stable state. Spotted Seatrout did the same, but it takes almost ten times as long to see that trend due to the smaller rd value. Gizzard Shad decrease to zero similarly to how they do with a density-independent model, but reaching zero takes much less time with the density-dependent model.

Because Black Drum and Spotted Seatrout reached their carrying capacitites, density-dependent effects are likely important, meaning a density-dependent model will work best to project the populations of these two species. Conversely, because Gizzard Shad decreased to zero (away from K) in both cases, a density-independent model works to assess the population growth of this species.

## 1.2 Based on the selected models, how does lambda change over time?

```{r}

# Density-independent lambda over time

sim_with_lambda <- sim |>
  arrange(species, year) |>
  group_by(species) |>
  mutate(
    lambda_realized = lead(projection) / projection
  ) |>
  ungroup()

ggplot(sim_with_lambda |> filter(!is.na(lambda_realized), year <= 100),
       aes(x = year, y = lambda_realized, color = species)) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "black") +
  facet_wrap(~species, scales = "free_y", ncol = 1) +
  labs(
    title = "Lambda Over Time, Density-Independent Model",
    x = "Years",
    y = "Lambda"
  ) +
  theme_bw()
  

# Density-dependent lambda over time

sim2_with_lambda <- sim2 |>
  arrange(species, year) |>
  group_by(species) |>
  mutate(
    lambda_realized = lead(projection) / projection
  ) |>
  ungroup()

ggplot(sim2_with_lambda |> filter(!is.na(lambda_realized), year <= 1000),
       aes(x = year, y = lambda_realized, color = species)) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "black") +
  facet_wrap(~species, scales = "free_y", ncol = 1) +
  labs(
    title = "Lambda Over Time, Density-Dependent Model",
    x = "Years",
    y = "Lambda"
  ) +
  theme_bw()
  
```

In a density-independent model, lambda is constant. However, in a density-dependent model, the trends in lambda decreased in a sigmoidal fashion towards 1 for Black Drum and Spotted Seatrout, while for Gizzard Shad, lambda started below one and decreased exponentially until leveling out.

## 2.1 Does the sampling assigned to your group reach the asymptote of the species accumulation curve? Discuss results.

```{r}
Rare <- data |> 
  group_by(date) |> 
  ungroup()
sp_rare  <- unique(Rare$species)
yr_rare <- unique(data$date)
sp_rare_tib = tibble(n_samp = 1:length(yr_rare), n_spp = NA)
for (i in 1:length(yr_rare)){
  # sample ID to include
  samp = yr_rare[1:i]
  # include only sample numbers 
  d = data |> 
    filter(date %in% samp,
           n > 0)
  sp_rare_tib$n_spp[i] = length(unique(d$species))
}
ggplot(sp_rare_tib, aes(n_samp, n_spp))+
  geom_line(linewidth = 1)+
  labs(x = 'Number of Samples',
       y = 'Number of Species')+
  theme_bw()
```

(Answer)

## 2.2 Using the species that composed 75% of the abundance of species within the nektonic community, assess the annual trends of alpha and beta diversity, and the inverse Simpson diversity index.

```{r}
data2 <- data |>
  group_by(species) |>
  summarise(sum(n))
sum(data2$`sum(n)`)
16745*0.75
sp75 <- c('Black Drum', 'Red Drum', 'Atlantic Croaker', 'Blue Crab', 'Striped Mullet')
data <- data %>%
  mutate(date = mdy(date))
data <- data %>%
  mutate(year = year(date))
annual_abundance <- data %>%
  group_by(species, year) %>%
  summarise(total_abundance = sum(n, na.rm = TRUE), .groups = "drop")
annual_abundance <- annual_abundance |>
  filter(species %in% sp75)

df = unique(annual_abundance[c("year")])
df$H = NA
annual_abundance <- annual_abundance %>%
  rename(abundance = total_abundance)
for (i in 1:nrow(df)){
  d = annual_abundance |> filter(year == df$year[i])
  d = d |> 
    count(species,wt = abundance) |> 
    mutate(pi = n/sum(n),
           ln_pi = log(pi),
           p_ln_pi = pi*ln_pi)
  
  df$H[i] = -sum(d$p_ln_pi)
}
df |>
  summarise(mean_H = mean(H, na.rm = TRUE),
            sd_H = sd(H, na.rm = TRUE))
ggplot(df, aes(year, H)) +
  geom_line(linewidth = 1)+ 
  geom_point() +
  labs(x = 'Date',
       y = 'Shannon Diversity')+
  theme_bw()

df$d <- NA
for (i in 1:nrow(df)){
  d = annual_abundance |> filter(year == df$year[i])
  d = d |> count(species,wt = abundance) |> 
    mutate(pi = n/sum(n))
  
  df$d[i] = sum(d$pi^2)
}

df$even <- NA
df$inv <- NA
for(i in 1:nrow(df)){
df$even[i] = 1 - df$d[i]
df$inv[i] = 1/df$d[i]}

ggplot(df, aes(year, d)) +
  geom_line(linewidth = 1)+ 
  geom_point() +
  labs(x = 'Date',
       y = 'Simpson Diversity')+
  theme_bw()

ggplot(df, aes(year, inv)) +
  geom_line(linewidth = 1)+ 
  geom_point() +
  labs(x = 'Date',
       y = 'Inverse Simpson Diversity')+
  theme_bw()

mean(df$inv)

data_w <- data |>
  group_by(year, species) |>            
  summarise(total_n = sum(n), .groups="drop") |>
  pivot_wider(names_from = species, values_from = total_n, values_fill = 0)
comm <- data_w |> 
  summarise(across(`Black Drum`:`White Shrimp`, mean))

jacdist <- vegdist(data_w, method = "jaccard")
nmds.jc <- metaMDS(data_w, distance = "jaccard", k = 2, try = 100)
nmds.jc
braydist = vegdist(data_w, method = "bray")
braydist
nmds.bc = metaMDS(data_w, distance = "bray", k = 2, try = 100)
nmds_output = bind_rows(bc = data.frame(nmds.bc[["points"]]),
                        jc = data.frame(nmds.jc[["points"]])) |> 
  mutate(year = rep(unique(data_w$year), times = 2),
         Dissimilarity = rep(c("Bray", "Jaccard"),
                             each = length(unique(year))))

ggplot(nmds_output, aes(MDS1, MDS2, color = as.factor(year)))+
  geom_point(size = 4)+ 
  facet_wrap(~Dissimilarity)+
  labs(color = 'year')+
  scale_color_viridis_d(option = 'turbo')+
  theme_bw()
```

## 2.2 Using a community trajectory analysis, assess the annual stability of the fish community across the sites from the assigned basin and gear type. Describe and discuss the main results.

```{r}
data$site <- as.integer(factor(data$lat))
data_w2 <- data |>
  pivot_wider(names_from = species,
    values_from = n,
    values_fill = 0)
comm_matrix <- data_w2 |> select(-year, -basin, -date, -lat, -lon, -gear, -sci_name, -site)
site <- data$site
survey = data$year
b_dist <- vegdist(comm_matrix, method = "bray", na.rm = TRUE)
d <- defineTrajectories(b_dist, sites = site, surveys = survey)
trajectoryPCoA(d,
  traj.colors = c("black","red", "blue", 'lightblue3', 'purple', 'pink'),
  lwd = 3,
  survey.labels = FALSE)
legend("bottomleft", col=c("black","red", "blue", 'lightblue3', 'purple', 'pink'), 
       legend=unique(data$site), bty="n", lty=1, lwd = 2)
```

(Answer)
